---
title: "Homework 3"
subtitle: "Time Series Graphics Part 2"
eval: false
---

```{r}
#| eval: true
#| include: false
library(tidyverse)
library(fpp3)
```

# Getting started

### Start with the following steps:

-   Go to our RStudio Server at <http://turing.cornellcollege.edu:8787/>

-   Open the respective file from the shared STA364_inst_files folder mentioned above. It will be named something like *hw-03_LAST_NAME.qmd*.

-   Then you need to save your copy. Click *File -\> Save as -\> Navigate to the folder `STA364_Projects` (that we share) -\> Change the "LAST_NAME part of the file name to your last name -\> Save*.

-   Update the top of the document, called the `YAML,` with your name.

# Homework Instructions

::: callout-note
Be sure to include the relevant R code and complete sentences answering each question (e.g., if I ask for the average, you can output the answer in R but **also** write a complete sentence with the answer). Be sure to save your files frequently!

From this point forward, you need to start commenting on your graphs. What do you observe? Are there trends/cycles/seasonal effects? Outliers? Other interesting features?
:::

Data for the homework is in the *STA364_inst_files \> data* folder.

# Exercises

### Question 1 (fpp7)

1.  Monthly Australian retail data is provided in `aus_retail`. Select one of the time series as follows (but choose a different seed value):

```{r q1}
set.seed(12345678)
myseries <- aus_retail |>
  filter(`Series ID` == sample(aus_retail$`Series ID`,1))
```

Explore your chosen retail time series using the following functions: `autoplot`, `gg_season()`, `gg_subseries()`, `gg_lag()`, `ACF()|>autoplot()`.

### Question 2 (fpp8)

2.  Use the following graphics functions: `autoplot()`, `gg_season()`, `gg_subseries()`, `gg_lag()`, `ACF()` and explore features from the following time series: “Total Private” Employed from *us_employment*, Bricks from *aus_production*, Hare from *pelt*, “H02” Cost from *PBS*, and Barrels from *us_gasoline*.

<!-- -->

a.  Can you spot any seasonality, cyclicity, or trend?

b.  What do you learn about the series?

c.  What can you say about the seasonal patterns?

d.  Can you identify any unusual years?

### Question 3 (fpp10)

3.  The *aus_livestock* data contains the monthly total number of pigs slaughtered in Victoria, Australia, from Jul 1972 to Dec 2018. Use `filter()` to extract pig slaughters in Victoria between 1990 and 1995. Use `autoplot()` and `ACF()` for this data. How do they differ from white noise? If a more extended period of data is used, what difference does it make to the ACF?

### Question 4 (fpp11)

4.  

    a.  Use the following code to compute the daily changes in Google closing stock prices. Explain what lines 2 through 5 are doing.

```{r q4}
dgoog <- gafa_stock |>                                    # line 1
  filter(Symbol == "GOOG", year(Date) >= 2018) |>         # line 2
  mutate(trading_day = row_number()) |>                   # line 3
  update_tsibble(index = trading_day, regular = TRUE) |>  # line 4
  mutate(diff = difference(Close))                        # line 5

```

b.  Why was it necessary to re-index the tsibble?

c.  Plot these differences and their ACF.

d.  Do the changes in the stock prices look like white noise?

# Submission

When you are finished with your homework, be sure to Render the final document. Once rendered, you can download your file by:

-   Finding the .html file in your File pane of RStudio (on the bottom right of the screen)
-   Click the check box next to the file
-   Click the blue gear above and then click "Export" to download
-   Submit your final HTML document to the respective assignment on [Moodle]({{< var course.moodle >}})
