---
title: "Homework 2"
subtitle: "Time Series Graphics"
eval: false
---

```{r}
#| eval: true
#| include: false
library(tidyverse)
library(fpp3)
```

The goal of this assignment is to introduce you to R, RStudio, and Time Series!

# Getting started

## R/Rstudio in Class

-   Go to our RStudio Server at <http://turing.cornellcollege.edu:8787/>

### Each of your assignments will begin with the following setup steps.

-   Finding the instructions on our website: <https://stats-tgeorge.github.io/STA363_AdvReg/>

-   Going to our RStudio Server at <http://turing.cornellcollege.edu:8787/>

-   Open the respective file from the shared STA364_inst_files folder mentioned above. It will be named something like *hw-02_LAST_NAME.qmd*.

-   Then you need to save your copy. Click *File -\> Save as -\> Navigate to the folder `STA364_Projects` (that we share) -\> Change the "LAST_NAME part of the file name to your last name -> Save*.

-   Update the top of the document, called the `YAML` with your name. 

# Homework Instructions

Be sure to include the relevant R code and full sentences answering each of the questions (e.g., if I ask for the average, you can output the answer in R but **also** write a full sentence with the answer). Be sure to save your files frequently!

Data for the homework will be in the STA364_inst_files -\> data folder.

# Exercises

## Reading

Read [fpp3](https://otexts.com/fpp3/what-can-be-forecast.html) 2.7-2.9.

## Problems

### Question 1

Explore the following four time series: *Bricks* from `aus_production`, *Lynx* from `pelt`, *Close* from `gafa_stock`, *Demand* from `vic_elec`.

a. Use ? (or help()) to find out about the data in each series.
b. What is the time interval of each series?
c. Use `autoplot()` to produce a time plot of each series.
d. For the last plot, modify the axis labels and title. See the `labs()` function. 


### Question 2 (fpp3)

i. Create a data folder inside your STA364 project folder. Go to RStudio and navigate to your STA364Project folder in the *files* tab on the bottom right. Click *New Folder* and name the folder *data*. 

ii. Now you need to copy the data from the instructor data folder to a data folder in your own folder. Go to the STA364_inst_files folder, and into the data folder 

iii. Check the box next to `tute.csv`, then *More*, then *Copy to...*; save it in your newly created data folder.  

The first chunk below reads `tute1.csv` into R and use then shows the data with the `View()` function. You should find four columns of information. Columns B through D each contain a quarterly series, labelled *Sales*, *AdBudget* and *GDP*. *Sales* contains the quarterly sales for a small company over the period 1981-2005. *AdBudget* is the advertising budget and *GDP* is the gross domestic product. All series have been adjusted for inflation.

You can read the data into R with the following script:

```{r}
tute1 <- readr::read_csv("data/tute1.csv")
View(tute1)
```

Then converts the data to a time series object in R (tsibble).

```{r}
mytimeseries <- tute1 |>
  mutate(Quarter = yearquarter(Quarter)) |>
  as_tsibble(index = Quarter)
```

Then construct the time series plot for the *Sales* series.

```{r}
mytimeseries |>
  autoplot(Sales)
```

a. Use this code to generate a time series plot for the other two series in the data, *AdBudget*, and *GDP*. 
b. Compare and contrast the three series. 

### Question 3 (fpp4)

The `USgas` package contains data on the demand for natural gas in the US.

a. Install the `USgas` package. (Use `install.packages('package_name')`). Then to use it you will need `library(package_name)`.
b. Create a tsibble from `us_total` with *year* as the index and *state* as the key.
c. Plot the annual natural gas consumption by state for the New England area (comprising the states of Maine, Vermont, New Hampshire, Massachusetts, Connecticut and Rhode Island).


### Question 4 (fpp6)

The `aus_arrivals` data set comprises quarterly international arrivals to Australia from Japan, New Zealand, UK and the US. The data is available after you have called in the `fpp3` package. 

a. Use `autoplot()`, `gg_season()` and `gg_subseries()` to compare the differences between the arrivals from these four countries. 

b. Can you identify any unusual observations?



# Submission

When you are finished with your homework, be sure to Render the final document. Once rendered, you can download your file by:

-   Finding the .html file in your File pane of RStudio (on the bottom right of the screen)
-   Click the check box next to the file
-   Click the blue gear above and then click "Export" to download
-   Submit your final HTML document to the respective assignment on [Moodle]({{< var course.moodle >}})


