---
title: "Homework 3"
author: "REPLACE WITH NAME"
format:
  html:
    embed-resources: true
---

### Question 1 (fpp7)

1.  Monthly Australian retail data is provided in `aus_retail`. Select one of the time series as follows (but choose a different seed value):

```{r q1}
set.seed(12345678)
myseries <- aus_retail |>
  filter(`Series ID` == sample(aus_retail$`Series ID`,1))
```

Explore your chosen retail time series using the following functions: `autoplot`, `gg_season()`, `gg_subseries()`, `gg_lag()`, `ACF()|>autoplot()`.

```{r q1}

```

TYPE ANSWER


### Question 2 (fpp8)

2.  Use the following graphics functions: `autoplot()`, `gg_season()`, `gg_subseries()`, `gg_lag()`, `ACF()` and explore features from the following time series: “Total Private” Employed from *us_employment*, Bricks from *aus_production*, Hare from *pelt*, “H02” Cost from *PBS*, and Barrels from *us_gasoline*.

a.  Can you spot any seasonality, cyclicity, or trend?

b.  What do you learn about the series?

c.  What can you say about the seasonal patterns?

d.  Can you identify any unusual years?


#### Total Private from *us_employment*
```{r q2_employment}

```

TYPE ANSWER

#### Bricks from *aus_production*

```{r q2_production}

```


TYPE ANSWER

#### Hare from *pelt*

```{r q2_pelt}

```


TYPE ANSWER

#### H02 from *PBS*

```{r q2_PBS}

```


TYPE ANSWER

#### Barrels from *us_gasoline*

```{r q2_gasoline}

```


TYPE ANSWER

### Question 3 (fpp10)

3.  The *aus_livestock* data contains the monthly total number of pigs slaughtered in Victoria, Australia, from Jul 1972 to Dec 2018. Use `filter()` to extract pig slaughters in Victoria between 1990 and 1995. Use `autoplot()` and `ACF()` for this data. How do they differ from white noise? If a more extended period of data is used, what difference does it make to the ACF?


```{r q3}

```


TYPE ANSWER

### Question 4 (fpp11)

4.  

a.  Use the following code to compute the daily changes in Google closing stock prices. Explain what lines 2 through 5 are doing.

```{r q4}
dgoog <- gafa_stock |>                                    # line 1
  filter(Symbol == "GOOG", year(Date) >= 2018) |>         # line 2
  mutate(trading_day = row_number()) |>                   # line 3
  update_tsibble(index = trading_day, regular = TRUE) |>  # line 4
  mutate(diff = difference(Close))                        # line 5

```

b.  Why was it necessary to re-index the tsibble?

TYPE ANSWER

c.  Plot these differences and their ACF.

TYPE ANSWER

d.  Do the changes in the stock prices look like white noise?

TYPE ANSWER