---
title: "Homework 2"
author: "REPLACE WITH NAME"
format:
  html:
    embed-resources: true
---

### Question 1

Explore the following four time series: *Bricks* from `aus_production`, *Lynx* from `pelt`, *Close* from `gafa_stock`, *Demand* from `vic_elec`.

a. Use ? (or help()) to find out about the data in each series.

```{r help}

```

b. What is the time interval of each series?

WRITE YOUR ANSWER HERE

c. Use `autoplot()` to produce a time plot of each series.

```{r brick_plot}

```

```{r lynx_plot}

```

```{r close_plot}

```

```{r demand_plot}

```

d. For the last plot, modify the axis labels and title. See the `labs()` function. 




### Question 2 (fpp3)


The first chunk below reads `tute1.csv` into R and use then shows the data with the `View()` function. You should find four columns of information. Columns B through D each contain a quarterly series, labelled *Sales*, *AdBudget* and *GDP*. *Sales* contains the quarterly sales for a small company over the period 1981-2005. *AdBudget* is the advertising budget and *GDP* is the gross domestic product. All series have been adjusted for inflation.

You can read the data into R with the following script:

```{r}
tute1 <- readr::read_csv("data/tute1.csv")
View(tute1)
```

Then converts the data to a time series object in R (tsibble).

```{r}
mytimeseries <- tute1 |>
  mutate(Quarter = yearquarter(Quarter)) |>
  as_tsibble(index = Quarter)
```

Then construct the time series plot for the *Sales* series.

```{r}
mytimeseries |>
  autoplot(Sales)
```

a. Use this code to generate a time series plot for the other two series in the data, *AdBudget*, and *GDP*. 

```{r, ts_AdBudget}

```

```{r, ts_GDP}


```

b. Compare and contrast the three series. 


WRITE YOUR ANSWER HERE


### Question 3 (fpp4)

The `USgas` package contains data on the demand for natural gas in the US.

a. Install the `USgas` package. (Use `install.packages('package_name')`). Then to use it you will need `library(package_name)`.

```{r pkg_install}
#| eval: false # you do not need to install it every render
# this will show the code but since eval (evaluation) is false,
# it will run the chunk at each render


```


b. Create a tsibble from `us_total` with *year* as the index and *state* as the key.

```{r 3b}


```

c. Plot the annual natural gas consumption by state for the New England area (comprising the states of Maine, Vermont, New Hampshire, Massachusetts, Connecticut and Rhode Island). Interpret your results. 

```{r 3c}

```



### Question 4 (fpp6)

The `aus_arrivals` data set comprises quarterly international arrivals to Australia from Japan, New Zealand, UK and the US. The data is available after you have called in the `fpp3` package. 

a. Use `autoplot()`, `gg_season()` and `gg_subseries()` to compare the differences between the arrivals from these four countries. 

```{r 4a_J}

```

```{r 4a_NZ}

```

```{r 4a_UK}

```

```{r 4a_US}

```

b. Can you identify any unusual observations?

WRITE YOUR ANSWER HERE




